A Domain-Specific Optimizing Compiler\\for Solving Partial Differential Equations

### 2014-02-18 paper/bare_conf.tex
Add explanation about why we do not show blas results

\section{Introduction}
In many fields, like computational fluid dynamics, computational electromagnetics, and structural mechanics, programs model phenomena by means of partial differential equations (PDEs). Numerical techniques, like Finite Volume Method (FVM) and Finite Element Method (FEM), are widely employed in real-world applications to approximate solutions of PDEs. Unstructured meshes are often used to discretize the domain of a PDE, since their geometric flexibility allows solvers to be extrimely effective. However, they are also one of the main responsible for performance degradation, due to the introduction of indirect accesses (e.g. \texttt{A[B[i]]}) to access mesh components. A typical problem, therefore, is that implementing a scientific application requires a great effort, other than domain knowledge, and even more time is needed to optimize their execution time.

@@

OP2, as well as Lizst~\cite{lizst}, adopt a kernel-oriented programming model, in which the computation semantics is expressed through self-contained functions (or ``kernels''). A kernel is applied to all elements in a set of mesh components (e.g. edges, vertices, elements), with an implicit synchronization between the application of two consecutive kernels. On commodity multi-cores, a kernel is executed sequentially by a thread, while parallelism is achieved partitioning the mesh and assigning each partition to a thread. Together with the indirections problem, kernel optimization is the major concern in unstructured mesh applications. In this paper, we tackle this problem by proposing a domain-driven optimization strategy for a class of kernels used in FEM.

@@

The compiler we developed is integrated with PyOP2, and already used in the release of Firedrake~\cite{firedrake-code}. This allows us to evaluate the proposed code transformations in a range of real-world problems, employing different low-order methods (from $p=1$ to $p=4$; the higher, the more expensive a kernel is, but better is the accuracy), and varying the topology of the mesh (2D, 3D, 3D hybrid structured-unstructured, which in turn affect the kernel cost). Performance results show that compiler-generated code for non-trivial assembly kernels, if not adequately transformed, is sub-optimal. A cost-model-driven sequence of source-to-source code transformations, aimed at improving register allocation and register data locality, can result in performance improvements up to 33$\%$ over ``softly-optimized'' code (i.e. where only basic transformations are performed, such as auto-vectorizable loop-invariant code motion, padding and data alignment), and up to 77$\%$ over the original FFC kernels. The contribution of this paper is therefore twofold:
\begin{itemize}
\item An optimisation strategy for a class of kernels widely-used in scientific applications, namely Finite Element Local Assembly kernels. Our approach exploits domain knowledge to go beyond the limits of both vendor (e.g. $icc$) and research (e.g. those based on the polyhedral model) compilers.
\item Automation of such code optimizations in Firedrake, a framework for the resoloution of PDEs through FEMs. This enables performing an in-depth performance evaluation of the proposed optimization strategy in real-world simulations.
\end{itemize}

@@

%the focus of optimization should be on managing registers, exploiting ILP in its various forms, and reducing loop overhead

@@

%What we do in computational science. Applications motivating our study. The programming model of lizst and op2: ``simple'' kernels applied to sets of mesh elements. The need for optimizing such kernels. Contributions of the paper:
%\begin{itemize}
%\item A \textbf{compiler-driven} optimisation approach that can be generalised to a wide variety of kernels employed in many computational science codes. The key feature common to these kernels is that they usually fit the L1 cache of commodity processors.
%\item Automation of such code optimizations in a framework for the resoloution of partial differential equations through the Finite Element Method, called Firedrake. This allows us to perform an in-depth performance evaluation of the proposed optimization strategy in real-world computations.
%\end{itemize}

@@

The complexity of FFC-generated kernels depends on the mathematical problem being solved. In simpler cases, the loop nest, always characterized by affine 1-stride memory accesses, is perfect, it has short trip counts (in the range 3-15), and the computation reduces to a summation of a few, small outer-product vector-operations. An example is provided in Figure~\ref{code:helmholtz}, which shows an assembly kernel for a Helmholtz problem, using Lagrange basis functions on 2D elements with polynomial order $p=1$. The outer-product along the $j$ and $k$ dimensions computes the stiffness matrix $A$. In other scenarios, for instance when solving a non-linear problem like Burgers (see Figure~\ref{code:burgers}), the number of arrays involved in the computation of $A$ can be much larger: in this case, 14 unique arrays are accessed, and the same array can be referenced multiple times within the expression. Also, constants evaluated in outer loops (called $F$ in the code), acting as scaling factors of arrays, may be required; trip counts can be larger (proportionally to the order of the method); arrays may be block-sparse. Despite such a big space of possible assembly implementations, it is still possible to identify a domain-specific structure common to all kernels, which can be turned into effective code transformations and simd-vectorization.

\section{A Domain-Driven Approach to Code Optimization}
%All the work done in the polyhedral model should be useful, but our kernels fit L1, so they need special attention for what concerns key aspects like managing registers, exploiting ILP in its various forms, and reducing loop overhead. 
%Follows a description of code transformations using our compiler-driven approach.

@@

\section{Performance Evaluation}
%Performance evaluation. We show all of the results got so far and we explain them. We demonstrate the claim that a compiler-driven approach is necessary to obtain maximum performance. Also, our results prove that current vendor compilers still have notable limitations, despite optimizing loop nests that are 1) affine, 2) have no stencil, and 3) ?
\begin{figure}[h]
\centerline{\includegraphics[scale=0.7]{Pictures/helmholtz-normalized.pdf}
\label{fig_first_case}}
\caption{Helmholtz}
\end{figure}

>>>>

\section{Introduction}
In many fields, like computational fluid dynamics, computational electromagnetics, and structural mechanics, programs model phenomena by means of partial differential equations (PDEs). Numerical techniques, like Finite Volume Method (FVM) and Finite Element Method (FEM), are widely employed in real-world applications to approximate solutions of PDEs. Unstructured meshes are often used to discretize the domain of a PDE, since their geometric flexibility allows solvers to be extrimely effective. However, they are also one of the main responsible for performance degradation, due to the need for indirect accesses (e.g. \texttt{A[B[i]]}) to access mesh components. A typical problem, therefore, is that implementing a scientific application requires a great effort, other than domain knowledge, and even more time is needed to optimize their execution time.

@@

OP2, as well as Lizst~\cite{lizst}, adopt a kernel-oriented programming model, in which the computation semantics is expressed through self-contained functions (or ``kernels''). A kernel is applied to all elements in a set of mesh components (e.g. edges, vertices, elements), with an implicit synchronization between the application of two consecutive kernels. On commodity multi-cores, a kernel is executed sequentially by a thread, while parallelism is achieved partitioning the mesh and assigning each partition to a thread. Together with the indirections problem, kernel optimization is one of the major concerns in unstructured mesh applications. In this paper, we tackle this problem by proposing a domain-driven optimization strategy for a class of kernels used in FEM.

@@

The compiler we have developed is integrated with PyOP2, and already used in the released version of Firedrake~\cite{firedrake-code}. This allows us to evaluate the proposed code transformations in a range of real-world problems, employing different low-order methods (from $p=1$ to $p=4$; the higher, the more expensive is a kernel, but better is the accuracy), and varying the topology of the mesh (2D, 3D, 3D hybrid structured-unstructured, which in turn affect the kernel cost). Performance results show that compiler-generated code for non-trivial assembly kernels, if not adequately transformed, is sub-optimal. A cost-model-driven sequence of source-to-source code transformations, aimed at improving register allocation and register data locality, can result in performance improvements up to 33$\%$ over ``softly-optimized'' code (i.e. where only basic transformations are performed, such as auto-vectorizable loop-invariant code motion, padding and data alignment), and up to 77$\%$ over the original FFC kernels. The contribution of this paper is therefore twofold:
\begin{itemize}
\item An optimisation strategy for a class of kernels widely-used in scientific applications, namely Finite Element Local Assembly kernels. Our approach exploits domain knowledge to go beyond the limits of both vendor (e.g. $icc$) and research (e.g. those based on the polyhedral model) compilers.
\item Automation of such code optimizations in Firedrake, a framework for the resoloution of PDEs through FEMs. This enables performing an in-depth performance evaluation of the proposed optimization strategy in real-world simulations.
\end{itemize}

@@

@@

@@

The complexity of FFC-generated kernels depends on the mathematical problem being solved. In simpler cases, the loop nest is perfect, it has short trip counts (in the range 3-15), and the computation reduces to a summation of a few, small outer-product vector-operations. An example is provided in Figure~\ref{code:helmholtz}, which shows an assembly kernel for a Helmholtz problem, using Lagrange basis functions on 2D elements with polynomial order $p=1$. In other scenarios, for instance when solving a non-linear problem like Burgers (see Figure~\ref{code:burgers}), the number of arrays involved in the computation of $A$ can be much larger: in this case, 14 unique arrays are accessed, and the same array can be referenced multiple times within the expression. Also, constants evaluated in outer loops (called $F$ in the code), acting as scaling factors of arrays, may be required; trip counts can be larger (proportionally to the order of the method); arrays may be block-sparse. Despite such a big space of possible assembly implementations, it is still possible to identify a domain-specific structure common to all kernels, which can be turned into effective code transformations and simd-vectorization.

The class of kernels we are considering has, in particular, some peculiarities. 1) The computation of the Jacobian, which is the first step of the assembly, is independent of the loop nest; this is not true in general, since the unstructured mesh might have bended elements that require the Jacobian be re-computed every $i$ iteration. 2) The stiffness matrix $A$ is always the result of an outer-product computation along the $j$ and $k$ dimensions. 3) Memory accesses along the three loop dimensions are always 1-stride .4) The $j$ and $k$ loops are interchangable, whereas to permute $i$ it might be necessary to precompute $F$ values and to introduce temporaries. 5) Basis functions arrays (denoted by $X$, $Y$, ...) are constants, and many sub-expressions on the right hand side of the stiffness matrix computation depend on just two loops (either $i$-$j$ or $i$-$k$). In Section~\ref{sec:code-transf} we show how to exploit these observations to define a set of systematic, composable optimizations.

One attractive possibility is to transform the loop nest into a sequence of calls to BLAS routines. There are several issues that make achieving this difficult and, potentially, not effective. The first problem concerns automating identification and extraction of matrix-matrix multiplications from the code, which requires a non-trivial analysis of the memory access pattern. Related to this problem is the fact that, in general, many sub-expressions should be pre-computed (i.e. computed once before the loop nest), which leads to using temporaries and, possibly, to working sets that do not fit the L1 cache (for $p>2$). In addition, using BLAS routines may impact data locality, since the same matrix can appear in multiple products, implying, at least, a certain degree of register spilling. Finally, it is known that BLAS routines perform far from peak performance when the dimension of the matrices is rather small~\cite{nek5000}. Hand-made implementations of the simple Helmholtz problem ($p=1,2,3,4$) as a sequence of BLAS ($dgemm$) routines have demonstrated that performance is worse than that shown in Section~\ref{sec:perf-results}. A comprehensive study concerning BLAS optimization is therefore left as further work.

\section{Code Transformations}
\label{sec:code-transf}
%All the work done in the polyhedral model should be useful, but our kernels fit L1, so they need special attention for what concerns key aspects like managing registers, exploiting ILP in its various forms, and reducing loop overhead. 
%Follows a description of code transformations using our compiler-driven approach.

@@

\section{Performance Evaluation}
\label{sec:perf-results}
%Performance evaluation. We show all of the results got so far and we explain them. We demonstrate the claim that a compiler-driven approach is necessary to obtain maximum performance. Also, our results prove that current vendor compilers still have notable limitations, despite optimizing loop nests that are 1) affine, 2) have no stencil, and 3) ?
\begin{figure}[h]
\centerline{\includegraphics[scale=0.7]{Pictures/helmholtz-normalized.pdf}
\label{fig_first_case}}
\caption{Helmholtz}
\end{figure}
